---
title: "Inclass Exercise 3"
author: "ZHU Fangyuan"
Naviga
editor: visual
---

## Overview

In this in-class exercise, we will learn several useful ways to deal with geospatial and aspatial data of **Nigeria** case! Let us get started now!

## 1 Packages Used

Packages below will be used in this in-class exercise:

-   ***sf*** : importing and processing geospatial data

-   ***tidyverse:*** importing and processing non-spatial data. In this exercise, readr package will be used for importing wkt data and dplyr package will be used to wrangling the data

```{r}
pacman::p_load(sf, tidyverse)
```

## 2 Data Used

In this exercise, two data sets will be used:

-   **Aspatial data:** from [WPdx Global Data Repositories](https://www.waterpointdata.org/access-data/). We will use the WPdx+ data set.

**Geospatial data:** Nigeria Level-2 Administrative Boundary (also known as Local Government Area) polygon features GIS data from the [Humanitarian Data Exchange](https://data.humdata.org/) portal or [geoBoundaries](https://www.geoboundaries.org/).

## 3 Data Preparation

### 3.1 Importing water point data

First, let us import the data used into the R environment using code below.

And the function `filter()` of ***dplyr*** will be used to filter data only belonging to Nigeria.

```{r}
wp_nga <- read_csv("aspatial/WPdx.csv") %>%
  filter(`#clean_country_name` == "Nigeria")
```

### 3.2 Convert wkt data

Notice that the newly imported tibble data frame (i.e. wp_nga) contains a field called *New Georeferenced Column* which represent spatial data in a textual format. In fact, this kind of text file is popularly known as **Well Known Text** in short **wkt**. So we need to convert the wkt data into **sf** data.frame.

First, `st_as_sfc()` of sf package is used to derive a new field called *Geometry* as shown in the code chunk below.

```{r}
wp_nga$Geometry = st_as_sfc(wp_nga$`New Georeferenced Column`)
```

Next, `st_sf()` will be used to convert the tibble data frame into sf data frame.

```{r}
wp_sf <- st_sf(wp_nga, crs=4326) 
wp_sf
```

### 3.3 Importing Nigeria LGA boundary data

Now we import the geospatial data in the R environment.

```{r}
nga<-st_read(dsn = "geospatial",
             layer = "geoBoundaries-NGA-ADM2",
             crs=4326)%>%
  select(shapeName)
```

### 3.4 Points in polygon

In this section, we will transfer the attribute information in *nga* sf data frame into *wp_sf* data frame.

```{r}
wp_sf <- st_join(wp_sf, nga)
write_rds(wp_sf,"wp_sf.rds")
write_rds(wp_nga,"wp_nga.rds")
```
