---
title: "Hands-on Exercise 3"
author: "Fangyuan ZHU"
editor: visual
---

# Geographical Segmentation with Spatially Constrained Clustering Techniques

## Overview

In this hands-on exercise, you will gain hands-on experience on how
to delineate homogeneous region by using geographically referenced
multivariate data. There are two major analysis, namely:

-   hierarchical cluster analysis

-   spatially constrained cluster analysis

## 1 Data used

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*):
    This is a GIS data in ESRI shapefile format. It consists of township
    boundary information of Myanmar. The spatial data are captured in
    polygon features.

-   *Shan-ICT.csv*: This is an extract of [The 2014 Myanmar Population and Housing Census Myanmar](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

## 2 Packages used

-    **sf, rgdal, spdep** for spatial data handling

-   **tidyverse, readr, ggplot2, dplyr** for attribute data handling

-   **tmap** for choropleth mapping

-   **coorplot, ggpubr, heatmaply** for multivariate data visualisation

-   **cluster, ClustGeo** for cluster analysis

Now let us install and launche these R packages into R environment!

```{r}
pacman::p_load(rgdal, spdep, tmap, sf, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse)
```

## 3 Data preparation

### 3.1 Importing data

#### 3.1.1 Importing geospatial data into R environment

In this section, you will import Myanmar Township Boundary GIS data and its associated attrbiute table into R environment by using *st_read()* of ***sf:***

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))
```

We can view the content of the newly created *shan_sf* simple features data.frame by using the code chunk below.

```{r}
shan_sf
```

Since the newly produced *shan_sf* is sf.data.frame which is conformed to tidy framework, we can also *glimpse()* to reveal the data type of its fields.

```{r}
glimpse(shan_sf)
```

#### 3,1,2 Importing aspatial data into R environment

Since the aspatial data is written in *csv* file, we use *read_csv()* to import the data into R environment and name it as *ict.*

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

It is saved in R\'s *tibble data.frame* format. We check the statistic attributes using *summary* function.
