---
title: "Inclass Ex2"
editor: visual
---

# Detection of clusters and outliers of Nigeria's water point

## Overview

## Preparation

### 1.1 Packages will be used

-   sf

-   tidyverse

-   tamp

-   spdep

-   funModeling

    ```{r}
    pacman::p_load(sf,tidyverse,tmap,spdep,funModeling)
    ```

### 1.2 Importing water point data

In this exercise, we will use two type of geospatial data. One is the boundary data of the Nigeria, one is the water point data.

```{r}
wp <- st_read(dsn = "geodata",
              layer = "geo_export",
              crs = 4326) %>%
  filter(clean_coun == "Nigeria")
```

Then, we save the the extracted data into the folder geodata.

```{r}
write_rds(wp,"geodata/wp_nga.rds")
```

### 1.3 Importing boundary data

```{r}
nga <- st_read(dsn = "geodata",
               layer = "geoBoundaries-NGA-ADM2",
               crs = 4326)
```

## 2. Data wrangling

### 2.1 Tackling with missing values

The raw data has missing values, so we should recode NA values in *status_cle* field into *Unknown* by using code chunk below.

```{r}
wp_nga <- read_rds("geodata/wp_nga.rds") %>%
  mutate(status_cle = replace_na(status_cle, "Unknown"))
```

### 2.2 Extracting water point values

We use EDA method to get the distribution of *status_cle field.*

```{r}
freq(data=wp_nga, 
     input = 'status_cle')
```

We use *filter()* to select functional classes in *status_cle* filed.

```{r}
wpt_functional <- wp_nga %>%
  filter(status_cle %in%
           c("Functional", 
             "Functional but not in use",
             "Functional but needs repair"))
write_rds(wpt_functional,"geodata/wpt_functional.rds")
```
